---
title: "Multiomics workshop part 3: Multi-omics integration with DIABLO"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{css, echo = FALSE}
.gt_table {
    margin-bottom: 20px;
    margin-top: 20px;
}
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Background

Up to now, we have been working with a single omics dataset at once. Although we can do multivariate analyses with a single omics dataset, we have greater ambitions. We want to know whether there are patterns in the data that cut across multiple omics and also are associated with our variable of interest (time, in this case study). 

## N-Integration 

N-Integration is the framework of having multiple datasets which measure different aspects of the same samples. For example, you may have transcriptomic, genetic and proteomic data for the same set of cells. N-integrative methods are built to use the information in all three of these dataframes simultaneously.

DIABLO is a novel `mixOmics` framework for the integration of multiple data sets while explaining their relationship with a categorical outcome variable. DIABLO stands for **D**ata **I**ntegration **A**nalysis for **B**iomarker discovery using **L**atent variable approaches for **O**mics studies. It can also be referred to as Multiblock (s)PLS-DA. 

## What is sPLS-DA?

DIABLO is based on a technique called sPLS-DA, which stands for *sparse partial least squares discriminant analysis*. But what is that? Let's start by explaining what partial least squares is and (vaguely) how it works.

### Partial least squares

Partial least squares is a statistical technique. Like many of the statistical techniques we are familiar with, starting with fitting a line through points in a scatter plot and going all the way up to the neural networks underlying AI large language models, it is a form of regression. The objective of PLS is to find the relationships between two matrices *X* and *Y*, by identifying the multidimensional direction in *X* that explains the maximum variance in the *Y* multidimensional space. Partial least squares discriminan analysis (PLS-DA) is just a variant of PLS where *Y* is a single categorical variable. We can use PLS-DA in this case because our response variable is just time point, with three discrete categories. The goal of PLS-DA is to summarize the original data in fewer dimensions, maximizing variation between groups and minimizing variation within groups.

In PLS, we typically choose a number of components, meaning the number of dimensions we want to reduce our many-dimensional dataset down to, while preserving as much information about it as possible. You can see in the diagram below that our original dataset *X* is of dimensions $N \times P$, where *N* is the number of samples and *P* the number of features. Usually in omics analyses, *P* is much bigger than *N*: in other words we have a lot more variables (genes, metabolites, microbiome taxa, etc., which could number up to the thousands) than we have independent samples (maybe in the 10-100 range). The goal of PLS is to preserve as much information as possible from the original $N \times P$ dataset using *k* components (*k* = 3 in the diagram). We estimate *Q* components, which are each *N*-length vectors, and *k* loading vectors, which are each of length *P*. The product of a $N \times k$ matrix times a $k \times P$ matrix is a matrix of dimensions $N \times P$. We estimate the values of the component and loading vectors to get as close as possible to the original data. If *P* is a really big number, and *k* is small, and our PLS model does a good job of recapturing the initial data, we have reduced the amount of data points needed to explain the pattern from the original $NP$ down to $Nk + kP = Q(N+P)$ which is tiny in comparison.

### Sparse partial least squares discriminant analysis

Sparse partial least squares (sPLS), which includes a special variant for categorical responses called sPLS-DA, is a form of PLS where only a subset of the variables in *X* are used to explain *Y*. This is useful both for explaining variation in the data and for predicting the response. It helps with explanation because it narrows down the pool of candidate variables, and it helps with prediction because it protects against overfitting.

sPLS-DA uses an algorithm called LASSO to reduce the number of variables from the dataset that are used to construct the loading vectors. In other words, the loading vectors that we estimated in PLS are now forced to have a lot of zeros in them. The amount of information we are using to summarize the information in the original dataset is even less.

### Multiblock sPLS-DA

Multiblock sPLS-DA, the technique used by DIABLO, extends sPLS-DA to more than one *X* dataset. Now we're trying to find a set of components that does two things: maximizes the correlation between our datasets $X_1, X_2, ... X_Q$ and, as before, explains the most variation in our categorical variable of interest *Y*.

**A NOTE OF CAUTION**: It's important to note that the analyses we are doing here are biologically naive. We know that there are complex biological processes going on within these cells that involve RNA transcripts, proteins, and various metabolites. We're not explicitly incorporating any knowledge about those processes into the models we are creating here. We are just mining the data for statistical associations between features across multiple multivariate datasets. I use the word "features" intentionally here; the analytical techniques we are using here are completely ignorant of whether those vectors of numbers represent RNA, protein, or anything else. To the model, they are just numbers. Thus, it is very important to keep in mind that this type of analysis is exploratory in nature. It can help draw attention to patterns existing in the data, that may indicate that certain biological processes are going on. We can use the patterns to speculate about the processes. Of course, these data do come from a manipulative experiment where cells were randomly assigned to treatments; so that does give us some basis to talk about causation when it comes to the treatment groups. But when it comes to the statistical associations between features in the different omics datasets, ultimately it is an exploratory/correlational analysis. We can use it to draw a lot of pretty pictures that are interesting to talk about, but it should be viewed as a jumping-off point for more detailed future studies into the specific biological processes behind them. Rant over!!!

# Setup

## Load packages

Load packages; as before, you may need to install packages from CRAN or Bioconductor if you are running this code on your own machine.

```{r load packages}
library(mixOmics)
library(Mus.musculus)
```

## Import pre-transformed data

These data have already been transformed to the $log_2$ scale, and features with low or no variance have been filtered out. This was done in the first two notebooks.

```{r data import}
transcriptomics_log2_z <- read.csv('data/transcriptomics_log2.csv', row.names = 1)

proteomics_log2_z <- read.csv('data/proteomics_log2.csv', row.names = 1)

metabolomics_log2_z <- read.csv('data/metabolomics_filtered_log2.csv', row.names = 1)
```

## Prepare data for integration

Put the three data frames into a list. The data were imported in a format where each row is a feature and each column is a sample, but mixOmics requires the opposite: samples in rows and features in columns. So, we use the `t()` function to transpose each of the data frames. Also, mixOmics doesn't play well with missing data, so we use the `na.omit()` function to get rid of any rows with even a single missing value. That's a very simplistic way of dealing with missing data; in practice you may want to play with multiple imputation or other methods for imputing missing data. In this case, we end up removing only a few hundred of the ~15,000 transcriptome features, and none of the proteome features, but we lose about half of the 171 metabolome features.

```{r construct data list}
data <- list(transcripts = t(na.omit(transcriptomics_log2_z)),
             metabolites = t(na.omit(metabolomics_log2_z)),
             proteins = t(na.omit(proteomics_log2_z)))
```

Check the dimensions of each of the three omics datasets. They all have 18 rows, as expected, one for each sample (6 each from 3 different time points), with differing numbers of columns.

```{r dimensions}
lapply(data, dim)
```

Check to make sure each of the three omics datasets have their rows in the same order. 

```{r check row names}
lapply(data, row.names)
```

The samples are all ordered in the same way in each dataset, in increasing order by time point. Create a vector `Y` to assign the samples to groups (time points in this case).

```{r group vector}
Y <- factor(rep(c('0h', '4h', '24h'), each = 6), levels = c('0h', '4h', '24h'))
```

One final setup step. As we did in part 1, let's use the Mus.musculus package, which contains mouse gene annotation information, to replace the numerical gene IDs with annotations where possible. Some tweaking of the code was needed to not have duplicate column names, which can cause problems later on.

```{r replace annotations}
genes <- select(
  Mus.musculus, 
  keys=colnames(data$transcripts), 
  columns=c("SYMBOL", "TXCHROM", "ENTREZID"),
  keytype="ENSEMBL"
)

genes <- genes[!duplicated(genes$ENSEMBL),]
colnames(data$transcripts)[!duplicated(genes$SYMBOL) & !is.na(genes$SYMBOL)] <- genes$SYMBOL[!duplicated(genes$SYMBOL) & !is.na(genes$SYMBOL)]
```


# Pairwise correlations between omics datasets

Before we do the full multiomics analysis, we need to look at the correlations between each pair of omics datasets. This will help us determine the design weights to use in the DIABLO analysis. What are those? Well, we have to assign a weight between 0 and 1 to each pair of omics datasets. The weight we choose depends on the goal of the analysis. Higher weights close to 1 prioritize a descriptive analysis that explains the maximum amount of correlation in the dataset. Lower weights give higher prediction accuracy but do not have as much explanatory power. If prediction is your goal, then you should use a small weight, for example 0.1. 

Instead of trying to guess what weight we should use, let's use a data-driven approach. We will do a quick partial least squares (PLS) analysis, with only one component, on each pair of omics datasets. Then calculate the correlation between the components associated with each of the datasets from the PLS.

```{r pairwise PLS}
pls1 <- pls(data[["transcripts"]], data[["metabolites"]], ncomp = 1)
pls2 <- pls(data[["transcripts"]], data[["proteins"]], ncomp = 1)
pls3 <- pls(data[["metabolites"]], data[["proteins"]], ncomp = 1)

# calculate correlation of transcripts and metabolites
print(cor(pls1$variates$X, pls1$variates$Y))
# calculate correlation of transcripts and proteins
print(cor(pls2$variates$X, pls2$variates$Y))
# calculate correlation of metabolites and proteins
print(cor(pls3$variates$X, pls3$variates$Y))
```

The pairwise correlations are all very high, at least 0.9. We can use 0.9 as the design weight in our DIABLO analysis.

# Tuning the DIABLO model

Create the design matrix with a diagonal of 0s and all off-diagonal entries being 0.9.

```{r design matrix}
design <- matrix(0.9, ncol = length(data), nrow = length(data), 
                dimnames = list(names(data), names(data)))
diag(design) <- 0 

design
```

Fit an initial DIABLO model using the `block.splsda()` function. We input our list of three omics datasets called `data` to the `X` argument of the function, and our categorical time point variable as `Y`. We specify `ncomp = 4` to fit up to four components, and we also pass the `design` matrix we created earlier.

```{r fit basic DIABLO model}
basic.diablo.model <- block.splsda(X = data, Y = Y, ncomp = 4, design = design)
```

The next step is to tune the model using cross-validation to determine how many components is optimal to explain the most variation between groups in `Y` without overfitting. Now, because we have only three groups in `Y`, it is unlikely that we will need more than two components &mdash; if you have two groups in your data, you should only need one component to separate them, and you'd need two components for three groups, three for four groups, etc. So, we could probably use common sense to say that two components are adequate without doing the cross-validation. But let's do it anyway to confirm that intuition.

The `perf()` function is used to tune the model. The arguments `validation = 'Mfold', folds = 3` tell it to do 3-fold cross-validation. What this does is divide the data up into three equal groups and fit the model three times, each time excluding the data from one group. Then, use the model that was fit to the other two groups to predict the classes of the observations in the excluded group. Repeat this process for 1-4 components, then repeat the entire thing 10 times (`nrepeat = 10`) to make sure we didn't get a fluke result from the way we randomly split up the data.

```{r tune basic DIABLO model}
perf.diablo = perf(basic.diablo.model, validation = 'Mfold', 
                   folds = 3, nrepeat = 10, seed = 1234)
```

The `plot()` method for the cross-validation output shows two types of error rate: overall (ER) and balanced (BER). Because we have exactly 6 samples at each time point, the data are perfectly balanced, so the ER and BER lines are identical here. If we had observational data that was imbalanced, they might differ. It also shows three types of distance metric that can be used for the prediction.

What does the plot show us? First of all, it shows that the error rate declines as the number of components increases. This makes sense because we are allowing the model to have more parameters, the more components are included. It also shows that the error rate is basically zero when we have $k = 2$ components. As we guessed beforehand, more than 2 is not necessary and will really only lead to overfitting. We also see that the metric `max.dist` is not very good when there's only one component, but the difference between distance metrics is negligible at 2 components and up.

```{r component number tuning plot}
plot(perf.diablo) # plot output of tuning
```

The `perf.diablo` object also tells us the optimal number of components it identified, using various methods. Here is one of them. Two of the three distance metrics identify 2 as the optimal number of components. Mahalanobis distance identifies 4 as optimal, but it performed less well than the other metrics at 2 components, so we'll stick with 2. Simpler is better.

```{r optimal component numbers}
perf.diablo$choice.ncomp$WeightedVote
```

Given the results of the first tuning, let's choose $k = 2$ components and use the `centroids.dist` metric because it always performed as well or better than the other metrics.

We will now do a second tuning step, this time to select a number of features to keep on each component. Remember that DIABLO is based on *sparse* PLS, meaning that only a subset of the original data is used. We are also using cross-validation for this part of the process, but this time we have to fit the model many more times because of all the different combinations of numbers of features we are trying out. We do the cross-validation for all possible combinations out of the list in `test.keepX`.

This takes ~5 minutes to run. Because it is a relatively small omics dataset, I did not bother to do anything to speed up the process. If you are working with a bigger dataset, consider looking into the `BiocParallel` package which will allow you to parallelize the computations so that multiple processors can work simultaneously, each fitting models with different combinations of numbers of features. 

In practice you also may want to increase `nrepeat` to at least 10 repetitions.

```{r tune number of features}
test.keepX <- list(
  transcripts = c(1, 2, 3, 4, 5, 10, 15, 20, 30, 40),
  metabolites = c(1, 2, 3, 4, 5, 10),
  proteins = c(1, 2, 3, 4, 5, 10, 15, 20, 30, 40)
)

tune.MULTI = tune.block.splsda(X = data, Y = Y, ncomp = 2, 
                              test.keepX = test.keepX, design = design,
                              validation = 'Mfold', folds = 3, nrepeat = 1,
                              dist = "centroids.dist", seed = 123)
```

How many features on each component gave the best predictive performance in cross-validation? We see that other than gene transcripts, very few variables are needed to capture most of the information in the original dataset. In particular, component 2 has only one variable from each omics dataset with a nonzero loading.

Note that in practice you might want to use more finely spaced numbers of features in the tuning grid. I used less finely spaced numbers to save time for the demo.

```{r optimal number of features}
list.keepX <- tune.MULTI$choice.keepX
list.keepX
```

# Final DIABLO model

To review, we have done the following to decide on the parameters of our final DIABLO model:

- PLS on each pair of omics datasets to decide what cross-correlation values to put in the design matrix (`design`)
- Cross-validation on an initial DIABLO model to decide how many components to include in the final model (`ncomp`)
- Another cross-validation after setting the optimal number of components, to decide how many features from each omics dataset will have nonzero loadings on each component in the final model (`list.keepX`)

Input all those values into the `block.splsda()` function to fit our final DIABLO model!

```{r fit final model}
final.diablo.model = block.splsda(X = data, Y = Y, ncomp = 2, 
                          keepX = list.keepX, design = design)
```

## Look at selected variables

Which variables from each of the omics datasets had non-zero loadings on the two components? The function `selectVar()` pulls these out. The selected variables have patterns of abundance that differ in a 


The selected variables can be extracted with the function selectVar(), for example in the mRNA block, as seen below. Note that the stability of selected variables can be extracted from the output of the perf() function.

```{r list selected variables}
selectVar(final.diablo.model, block = 'transcripts', comp = 1)$transcripts$name 
selectVar(final.diablo.model, block = 'metabolites', comp = 1)$metabolites$name 
selectVar(final.diablo.model, block = 'proteins', comp = 1)$proteins$name

selectVar(final.diablo.model, block = 'transcripts', comp = 2)$transcripts$name 
selectVar(final.diablo.model, block = 'metabolites', comp = 2)$metabolites$name 
selectVar(final.diablo.model, block = 'proteins', comp = 2)$proteins$name 
```

## Plots

### Diagnostic correlation plot

The first thing to do is a diagnostic correlation plot that confirms that the correlations between each pair of omics were maximized as specified by the design matrix. The correlations are high and positive which looks good.

```{r diagnostic correlation plot}
plotDiablo(final.diablo.model, ncomp = 1, col.per.group = c("salmon", "darkolivegreen3", "turquoise3"))
```

### Individual scatter plots and arrow plot

Next, we can plot scatter plots of how the individual samples are represented in the two-dimensional space formed by the two components. A separate plot is generated for each of the three omics (or blocks, as they are referred to in the mixOmics package). 

The plot here shows that across all three omics, component 1 separates the 4h time point from the other two. Despite only having a few variables with nonzero loadings, component 2 further separates the 0h and 24h time points.

```{r scatter plots}
plotIndiv(final.diablo.model, ind.names = FALSE, legend = TRUE, 
          title = 'DIABLO Sample Plots', col = c("salmon", "darkolivegreen3", "turquoise3"))
```

A so-called "arrow plot" shows how well the three omics agree with one another. This plot has the same information as the plot we just made by `plotIndiv()` but with a slightly different twist: each of the 18 samples is shown as a central point with three arrows radiating from it, each one pointing to the sample's coordinate on the two PLS components with respect to one of the three omics. If the arrows are short and the representations of each sample in the three different omics are close to one another, that means the omics datasets agree with each other closely. As we would expect given the high (> 0.9) correlations between each pair of omics, we see that the transcriptome, metabolome, and proteome agree well with each other here. Regardless of which omics dataset you look at, the clustering pattern of the time points is virtually identical.

```{r arrow plot}
plotArrow(final.diablo.model, ind.names = FALSE, legend = TRUE, 
          title = 'DIABLO', col = c("salmon", "darkolivegreen3", "turquoise3"))
```

### Correlation circle plot

This is a plot of the correlations between each of the selected variables and the components. Recall that 30, 4, and 2 variables were selected from transcriptome, metabolome, and proteome, respectively, for component 1, and only 1 from each of the omics for component 2. They are nearly all perfectly positively or negatively correlated with the component they were selected on. This is expected.

```{r correlation circle plot}
plotVar(final.diablo.model, var.names = FALSE, 
        style = 'lattice', legend = TRUE,
        pch = c(16, 17, 15), cex = c(0.8,0.8,0.8), 
        col = c("salmon", "darkolivegreen3", "turquoise3"))
```

### Circos plot

A Circos plot is a fairly common type of visualization that you see in omics integration studies. I am not sure how effective they are at conveying information about the data, because they are somewhat unintuitive. Let's start by making a Circos plot for component 1 only, as indicated by argument `comp = 1`.

Let's break down what we are looking at here.

- You can see a slice of the pie for each of the three omics. Most of the circle is taken up by the transcriptome data because 30 variables were selected from that dataset to have nonzero loadings in the final model, compared to 4 for metabolome and only 2 for proteome.
- Around the outside are lines indicating the relative abundance of each feature at the three time points (there's no labeled y-axis because it's all just relative, but the further from the center, the higher the abundance). Of the selected genes from the transcriptome dataset, a bit more than half were abundant at time 4h and less abundant at times 0h and 24h, and the rest showed the opposite pattern. This matches what we saw in the individual scatterplots: component 1 separates 4h from the other two time points, so it makes sense that the variables selected for that component are the ones where the 4h time point differs the most from the other two.
- The spiderweb-like pattern in the middle consists of lines connecting pairs of features that are (a) in two different omics and (b) correlated with one another. I set the argument `cutoff = 0.97` to display a line at the very strict cutoff of $|r| = 0.97$. Even with this strict cutoff, two of the metabolites and one of the proteins are shown as being correlated with many of the genes in the transcriptome. The metabolites and protein are both more abundant at the 4h time point, so they are connected with purple lines (indicating positive correlation) with the genes that are expressed more highly at 4h, and black lines (indicating negative correlation) to the genes that are not.

```{r circos plot, fig.height=8, fig.width=8}
circosPlot(final.diablo.model, cutoff = 0.97, line = TRUE, comp = 1,
           color.blocks= c("salmon", "darkolivegreen3", "turquoise3"),
           color.cor = c("mediumpurple1","grey20"), size.labels = 1,
           size.variables = 0.4, linkWidth = 0.3, showIntraLinks = FALSE, var.adj=-0.33)
```

Because only one variable from each omics was selected on component 2, that's too few to make a Circos plot for component 2 only. But if we remove the `comp = 1` argument we can make a Circos plot with all the variables selected on components 1 and 2. It's pretty similar to the first one.

```{r circos plot both components, fig.height=8, fig.width=8}
circosPlot(final.diablo.model, cutoff = 0.97, line = TRUE, 
           color.blocks= c("salmon", "darkolivegreen3", "turquoise3"),
           color.cor = c("mediumpurple1","grey20"), size.labels = 1,
           size.variables = 0.4, linkWidth = 0.3, showIntraLinks = FALSE, var.adj=-0.33)
```

### Loading weights plot

This plot shows the selected variables for each of the three omics, as a horizontal bar plot. The bar for each variable shows the group that has the highest absolute value loading on component 1. 
Looking at the transcriptome plot in the upper left, we see that the selected genes have either a positive or a negative loading on component 1. Because positive loadings on component 1 happen to be associated with high abundance at the 4h time point, the `plotLoadings()` function automatically colors the genes with the 4h color (green) if they have a positive loading on component 1. The genes with a negative loading are either colored pink or blue depending on whether the gene was more abundant at the 0h or 24h time point.

```{r loading weights, fig.height=8, fig.width=8}
plotLoadings(final.diablo.model, comp = 1, contrib = 'max', method = 'median', size.name = 0.6, legend.color = c("salmon", "darkolivegreen3", "turquoise3"))
```

### Heat map

This heat map has a similar interpretation to the one we made in a previous session. In the heat map, the colors represent standardized abundance values for the features. Each column is a feature that was selected to have nonzero loading in the final DIABLO model. Each row is one of the 18 samples. If we don't supply a `comp` argument to the heat map plotting function `cimDiablo()`, it will default to showing us component 1. So as with some of the previous visualizations we've seen, this heat map once again emphasizes that component 1 separates out the 4h time point from the other two. All features shown here are either low in abundance at 4h and high at 0h and 24h (the columns on the left), or the opposite (the columns on the right).

```{r heatmap}
#dev.off()
n_colors <- 1000  # Number of colors in the gradient
color_palette <- colorRampPalette(c("turquoise4", "turquoise3", "paleturquoise1", "lemonchiffon", "peachpuff", "salmon", "darkred"))(n_colors)
cimDiablo(final.diablo.model, margins = c(2, 18), size.legend = 1, color.blocks = c("salmon", "darkolivegreen3", "turquoise3"), color.Y = c("gray60", "gray90", "gray70"), color = color_palette)
#quartz.save("mixOmics_heatmap", type = "png", device = dev.cur(), dpi = 300)
```

## Performance of the model

We can now look at how well the model predicts. In other words, working "backward" from the omics data, can we predict what time point a particular observation of omics data comes from? This is once again done by cross-validation, where we purposefully hide some data from the model fitting process, fit the model to the remaining data, and predict what groups the held-out data points belong to.

We see that the error rate is low with one component (~14% overall) and the prediction performance becomes perfect when adding the second component.

```{r CV evaluation of final model performance}
perf.diablo = perf(final.diablo.model, validation = 'Mfold',
                   folds = 3, nrepeat = 10, 
                   dist = 'centroids.dist', seed = 321) 

perf.diablo$WeightedVote.error.rate
```

It may not make a lot of sense why we want to check this, because we already know the time point for all the data that we have, and in practice you would not have a sample of omics data without knowing what time point it came from. After all, our goal here is more exploring patterns in existing omics data, than training a model so that we can predict outcomes for future omics data we collect. But even if our goal is inference and understanding rather than prediction, it is reassuring if our model has good predictive performance. We can say that our model has learned something from the data, which means that we can look at outputs from the model and use them to understand the data. The data are from the real world and the model is not, it's only an abstraction of reality. So it's only useful if we can tie it back to reality in some way.

On that inspiring note, let's end our workshop on multiomics integration!
